Does personal relevance attenuate citizens' persuasion by government branding? Evidence from a survey experiment on air-pollution policy

### 2019-07-27 paper3_APPENDIX_revision_01.Rmd
Add files via upload

This section aims to assess the concern that participants did not accept the authenticity of the weak, fictional policy plan, and did not find it reliable, and as a results lost trust in the rest of the survey. To assess that risk, I took advantage of the fact that the order in which the strong and weak policies were presented was randomized. To reiterate, half of the participants saw a weak plan first, and thereafter a strong plan, and for the other half it was the opposite. Hence, I examined whether the the strong policy was evaluated differently when it was presented first (which means that participants at that point did not sea weak policy yet and did not question the reliability of the survey), then when it was presented second (after they already saw a fictional policy). This comparison, across the two policies, is graphically presented in the following plot. It shows that the differences between the means and distributions of the first and the second policy are relatively small and insignificant. The small increase between the first and second round may be explained be the fact that participants used the first (weak) policy as a benchmark, and therefore compared the second policy with the first one. Altogether, these test tend to indicate that participants' exposure to the fictional weak policy did not have a strong influence on their answers to the subsequent questions.

>>>>

This section aims to assess the concern that participants did not accept the authenticity of the weak, fictional policy plan, and did not find it reliable, and as a results lost trust in the rest of the survey. To assess that risk, I took advantage of the fact that the order in which the strong and weak policies were presented was randomized. To reiterate, half of the participants saw a weak plan first, and thereafter a strong plan, and for the other half it was the opposite. Hence, I examined whether the strong policy was evaluated differently when it was presented first (which means that participants at that point did not see a weak policy yet and did not question the reliability of the survey), then when it was presented second (after they already saw a fictional policy). This comparison, across the two policies, is graphically presented in the following plot. It shows that the differences between the means and distributions of the first and the second policy are relatively small and insignificant. The small increase between the first and second round may be explained be the fact that participants used the first (weak) policy as a benchmark, and therefore compared the second policy with the first one. Altogether, these test tend to indicate that participants' exposure to the fictional weak policy did not have a strong influence on their answers to the subsequent questions.

### 2019-07-27 paper3_revision_text.Rmd
Add files via upload

---
title: "Does personal relevance attenuate citizens' persuasion by government branding? Evidence from a survey experiment on air-pollution policy"
author: '**Saar Alon-Barkat**'
date: " "
always_allow_html: yes
output:
  word_document: default
  pdf_document:
    toc: yes
  html_document:
    theme: flatly
    toc: yes
    toc_float:
      collapsed: no
      smooth_scroll: yes
      toc_depth: 3
link-citations: yes
bibliography: phd_paper_2.bib
csl: https://raw.githubusercontent.com/citation-style-language/styles/master/apa.csl
urlcolor: blue
---

@@

*Recent studies demonstrated the potency of government branding to positively affect citizens' judgments of government organizations and policies. This study explores the boundaries of governments' persuasion of citizens through branding and symbolic communications. The latter is important given the the detrimental implications of the emotive effect of symbols, mainly its ability to compensate for organizations' poor functioning, and to elicit undue trust. Building on social psychology, I hypothesize that citizens are less susceptible to persuasion by symbolic communication, the more they perceive the policy issue in the communication as personally relevant. I test this expectation through a survey experiment, focused on air-pollution policy, utilizing natural variation in the perceived personal relevance between citizens residing in a polluted area and elsewhere. Contrary to my expectations, the results indicate that even high levels of perceived personal relevance do not attenuate the symbols' effect, which entails that the boundaries of persuasion through symbols are wider than expected.*

@@

The ability of government organizations to modify citizens' attitudes via such symbolic communications can be perceived favorably, insofar as it mitigates citizens' distrust in organizations and underevaluation of their performance. Accordingly, Symbolic communications, as opposed to mere factual information and substantive arguments, may enable organizations to overcome citizens' negative prior beliefs and biases against the public sector [@marvel_2015a; @marvel_2015; @hvidman_2016; @hvidman_2019]. Nonetheless, this form of persuasion is also highly problematic. Normatively, we expect citizens in a democratic society to form their opinions about their government based on their critical thinking, rather then on uncontious emotive responses. A yet more serious concern regards the potency of symbolic communication to compensate for organizations' poor performance, poorly planned policy plans and logically unpersuasive explanations, as indicated by previous studies [@alonbarkat_2017; @alonbarkat_2018]. Congruently, government branding can yield a misalignment between citizens' positive view and the "real world", which would undermine the democratic accountability and responsiveness of public organizations. Moreover, government agencies and/or politicians controlling them could use branding symbolic communication strategically as an instrument of propaganda, seeking to manipulate public opinion and mitigate justified public criticism of agencies' failures.

@@

[^footnote_other_paper]: The experiment was additionally designed to test a set of hypotheses regarding the interaction between symbols and information in communications. The methodological aspects which are relevant for examining the additional hypotheses are described in a separate publication. In this paper, I focus on those methodological and empirical parts of the study that are consequential for the examination of the moderating role or perceived personal relevance. A summary statistics and correlation of all variables, as well as the full survey are available in the supplementary Appendix, and replication materials are available in a github repository [url to be added upon publication].

@@

While comparing those residing in the Haifa Bay to others enables me to examine the moderating effect of perceived personal relevance, it still has an important limitation - it represents an observational comparison. The differences in personal relevance can be theoretically confounded by unobserved covariates, that cannot be effectively controlled in the analyses. I address this limitation in the research design in three main ways. *First*, in the selection of the research population, I sought to reduce differences in socioeconomic and demographic variables between residents of Haifa Bay, and their comparison group. Accordingly, I selected the major cities in the Haifa Bay area (*Haifa*, *Nesher*, *Qiryat-Hayiim*,[^footnote_qiryat-Hayiim] *Qiryat-Bialik*, *Qiryat-Mozkin*, *Qiryat-Ata* and *Qiryat-Yam*), and matched them with citizens from cities with a similar profile and population size at the center of the country (*Natania*, *Petach-Tikva* and *Rishon-Lezion*). The socioeconomic characteristics of these selected cities are summarized in the supplementary appendix (section ##). I further searched relevant EPM documents and news websites and confirmed that there no particular reporting about high air-pollution levels in the three center cities.

@@

[^footnote_unfiltered_sample]: In the supplementary appendix, I replicate the main models on the unfiltered sample, with no significant change to the results.

@@

Participants were then presented with the air-pollution and recycling policy plans (displayed in a random order). They were informed that these plans were obtained from EPM's annual work plan. The participants were randomly assigned to one of three conditions of symbolic elements in the communication (hereafter: symbols manipulation): treatment ("real symbols"), control-"no symbols" and control-"fake symbols". In the real symbols condition, the two policies are displayed with three symbolic brand elements, which have been used in EPM's public communications, and became widely recognized by the public. The first symbolic element is the unique brand logo of the EPM, which consists a pair of green and orange leaves that resemble two hands, designed to symbolize peace and harmony with the environment. The second element is the green color, which is strongly associated with taking care of the environment. The third element is an image of a celebrity comedian from an advertising campaign of the EPM. The Haifa Bay air-pollution policy (in the real symbols condition) was presented with an image from EPM campaign from 2010 under the slogan "starting to think green", starring the Israeli celebrity comedian Tal Friedman. The recycling policy (in the real symbols condition) was presented with an image from a campaign from 2017 for reducing the use of disposable bags ("taking every bag seriously"), starring the comedian Ido Rosenblum. In the control no-symbols condition, the two policies are displayed in a minimal black and white design, and without the logo and the celebrities’ images. Finally, in the control fake-symbols condition, the EPM logo is replaced by a fake logo, the green color is replaced with blue, and the images of celebrities from the campaigns are replaced with edited images of unfamiliar people. The purpose of the fake-logo condition was to distinguish the effect of the symbolic meaning of the EPM's graphic brand elements (i.e. the symbolic effect) from a possible effect of their aesthetic design. Accordingly, I specifically designed the "fake" symbols so that they would resemble the real symbols and have equivalent aesthetic qualities, without activating particular emotions and associations related to EPM, or environmental policy. The symbols conditions are displayed in **Appendix B**. Further details about the real and fake symbols are found in section ## of the supplementary appendix.

@@

[^footnote_reliability_information_manipulation]: A possible concern about that design is that participants would question the reliability of the weak policy. I assess that concern in the supplementary appendix (section ##).

@@

I further conducted power estimations for the main effects of the symbols and information manipulations, their conditional effects within each area and the aforementioned interactions. I conducted these estimations after the data collection and analyses, based on the analytical sample size. The estimated detectable (power=0.8) interaction coefficient between areas and the real symbols treatment is 0.27 SDs, and between areas and weak information manipulation is 0.23 SDs. As for the interactions with the relevance manipulation (among Haifa Bay respondents), I estimate that the detectable coefficients of their interactions with the symbols and information conditions are 0.36 and 0.29 SDs, respectively. The power estimations are reported in details in Section ## of the supplementary appendix.

@@

As for the relevance manipulation, data suggests that assignment to relevance priming questions did not increase Haifa Bay respondents' perceptions of personal relevance of the air-pollution policy. In fact, those assigned to the treatment group reported sightly lower level of personal relevance on average, yet these differences are not statistically significant (*t* = `r relevance.exp.t[["statistic"]] %>% round(3)`, *p*=`r round(relevance.exp.t[["p.value"]],3)`). This limitation should be kept in mind when interpreting the results regarding this manipulation. The results of all manipulation and validation checks are available from the supplementary appendix, along with summary statistics and correlation matrix for all variables.

@@

[^footnote_raw_means]: Means and SDs of outcome variables across the two areas and the experimental conditions are reported in section XX of the supplementary appendix.

@@

I further explored the robustness of these findings through additional analyses, reported in the supplementary appendix. *First*, I replicate the main models while restricting the analyses to the first policy seen by the participants. This enables to rule out the possibility that subjects' responses are affected by the manipulations in the first policy presented to them (a carryover effect), or by their evaluation of the it (assimilation or contrast effects). *Second*, I restrict the analyses to Haifa Bay residents, and compare between those who are more/less likely to perceive the air-pollution as personally relevant, based on two proxies: (a) Subjects city of residence, and its geographical proximity to the polluting industrial area; (b) parents for young children. *Third*, I replicated the analyses on the unfiltered sample. These additional analyses are consistent with the analyses presented in the paper, and do not provide evidence that greater levels of perceived personal relevance attenuates the effect of symbols.

@@

How can we explain the finding that the positive effect of the real symbols was stronger to some extent among Haifa Bay residents? Given that this pattern is also evident for the recycling policy, it is unlikely that it is linked to differences in perceived personal relevance. Rather, I tentatively suggest two other possible explanations for that. First, these patterns could be the result of the higher intercept of center residents' trust in EPM policies, or a ceiling effect. More generally, it could be the case that symbolic elements are more effective in shifting people from low to medium levels of trust (i.e. as in the case of Haifa Bay subjects), than shifting them from medium to high levels (i.e. as in the case of center subjects). In other words, symbols may be more effective in reducing citizens' distrust, then increasing their trust.[^footnote_ceiling_effect] A second explanation could be that EPM symbols attenuated the effect of Haifa Bay residents' prior beliefs on their judgments of the policy plans, an effect that can be related to motivated reasoning [@taber_2006; @christensen_2018; @baekgaard_2016]. It can be theorized that symbols attenuate motivated reasoning, when symbolic associations are not strongly related to identities of specific groups having strong prior positions on the policy or the organization (as in the case of the EPM symbols and the Haifa Bay locals). On the other hand, when symbolic associations do activate such group identities, they are likely to exacerbate partisan motivated reasoning in that regard, and polarization [@teodoro_2018]. For instance, it is possible that EPM "green" symbols enhance partisan motivated reasoning among groups divided on their support for environmental policy. Certainly, we must be cautious about these tentative explanations, which may be tested by future studies.

>>>>

---
title: "Does personal relevance attenuate citizens' persuasion by government branding? Evidence from a survey experiment on air-pollution policy"
author: '**Saar Alon-Barkat**'
date: " "
always_allow_html: yes
output:
  html_document:
    theme: flatly
    toc: yes
    toc_float:
      collapsed: no
      smooth_scroll: yes
      toc_depth: 3
  pdf_document:
    toc: yes
  word_document: default
link-citations: no
bibliography: phd_paper_2.bib
csl: https://raw.githubusercontent.com/citation-style-language/styles/master/apa.csl
urlcolor: blue
---

@@

*Recent studies have demonstrated the potency of government branding to positively affect citizens' judgments of government organizations and policies. Additionally, studies have pointed to the detrimental implications of the emotive effect of symbols, mainly its ability to compensate for organizations' poor functioning, and accordingly to elicit undue trust. In light of these concerns, this study explores the boundaries of governments' persuasion of citizens through branding and symbolic communications. Building on social psychology and marketing research, I hypothesize that citizens are less susceptible to persuasion by symbolic communication, the more they perceive the policy issue in the communication as personally relevant. I test this expectation through a survey experiment, focused on air-pollution policy in Israel, exploiting the natural variation in the perceived personal relevance between citizens residing in a polluted area in the country and others. The results consistently indicate that even high levels of perceived personal relevance do not significaly attenuate the effect of symbols, which entails that the boundaries of persuasion and manipulation through branding are wider than expected.*

@@

The ability of government organizations to modify citizens' attitudes via such symbolic communications can be perceived favorably, insofar as it mitigates citizens' distrust in organizations and underevaluation of their performance. Accordingly, Symbolic communications, as opposed to mere factual information and substantive arguments, may enable organizations to overcome citizens' negative prior beliefs and biases against the public sector [@marvel_2015a; @marvel_2015; @hvidman_2016; @hvidman_2019]. Nonetheless, this form of persuasion is also highly problematic. Normatively, we expect citizens in a democratic society to form their opinions about their government based on their critical thinking, rather then on uncontious emotive responses. A yet more serious concern regards the potency of symbolic communication to compensate for organizations' poor performance, poorly planned policy plans and logically unpersuasive explanations, as indicated by previous studies [@alonbarkat_2017] AUTHOR. Congruently, government branding can yield a misalignment between citizens' positive view and the "real world", which would undermine the democratic accountability and responsiveness of public organizations. Moreover, government agencies and/or politicians controlling them could use branding symbolic communication strategically as an instrument of propaganda, seeking to manipulate public opinion and mitigate justified public criticism of agencies' failures.

@@

[^footnote_other_paper]: The experiment was additionally designed to test a set of hypotheses regarding the interaction between symbols and information in communications. The methodological aspects which are relevant for examining the additional hypotheses are described in a separate publication. In this paper, I focus on those methodological and empirical parts of the study that are consequential for the examination of the moderating role or perceived personal relevance. A summary statistics and correlation of all variables, as well as the full survey are available in the supplementary Appendix (sections 5, 8), and replication materials are available in a github repository [url to be added upon publication].

@@

While comparing those residing in the Haifa Bay to others enables me to examine the moderating effect of perceived personal relevance, it still has an important limitation - it represents an observational comparison. The differences in personal relevance can be theoretically confounded by unobserved covariates, that cannot be effectively controlled in the analyses. I address this limitation in the research design in three main ways. *First*, in the selection of the research population, I sought to reduce differences in socioeconomic and demographic variables between residents of Haifa Bay, and their comparison group. Accordingly, I selected the major cities in the Haifa Bay area (*Haifa*, *Nesher*, *Qiryat-Hayiim*,[^footnote_qiryat-Hayiim] *Qiryat-Bialik*, *Qiryat-Mozkin*, *Qiryat-Ata* and *Qiryat-Yam*), and matched them with citizens from cities with a similar profile and population size at the center of the country (*Natania*, *Petach-Tikva* and *Rishon-Lezion*). The socioeconomic characteristics of these selected cities are summarized in the supplementary appendix (section 1). I further searched relevant EPM documents and news websites and confirmed that there no particular reporting about high air-pollution levels in the three center cities.

@@

[^footnote_unfiltered_sample]: In the supplementary appendix, I replicate the main models on the unfiltered sample, with no significant change to the results (section 7).

@@

Participants were then presented with the air-pollution and recycling policy plans (displayed in a random order). They were informed that these plans were obtained from EPM's annual work plan. The participants were randomly assigned to one of three conditions of symbolic elements in the communication (hereafter: symbols manipulation): treatment ("real symbols"), control-"no symbols" and control-"fake symbols". In the real symbols condition, the two policies are displayed with three symbolic brand elements, which have been used in EPM's public communications, and became widely recognized by the public. The first symbolic element is the unique brand logo of the EPM, which consists a pair of green and orange leaves that resemble two hands, designed to symbolize peace and harmony with the environment. The second element is the green color, which is strongly associated with taking care of the environment. The third element is an image of a celebrity comedian from an advertising campaign of the EPM. The Haifa Bay air-pollution policy (in the real symbols condition) was presented with an image from EPM campaign from 2010 under the slogan "starting to think green", starring the Israeli celebrity comedian Tal Friedman. The recycling policy (in the real symbols condition) was presented with an image from a campaign from 2017 for reducing the use of disposable bags ("taking every bag seriously"), starring the comedian Ido Rosenblum. In the control no-symbols condition, the two policies are displayed in a minimal black and white design, and without the logo and the celebrities’ images. Finally, in the control fake-symbols condition, the EPM logo is replaced by a fake logo, the green color is replaced with blue, and the images of celebrities from the campaigns are replaced with edited images of unfamiliar people. The purpose of the fake-logo condition was to distinguish the effect of the symbolic meaning of the EPM's graphic brand elements (i.e. the symbolic effect) from a possible effect of their aesthetic design. Accordingly, I specifically designed the "fake" symbols so that they would resemble the real symbols and have equivalent aesthetic qualities, without activating particular emotions and associations related to EPM, or environmental policy. The symbols conditions are displayed in **Appendix B**. Further details about the real and fake symbols are found the supplementary appendix (section 2).

@@

[^footnote_reliability_information_manipulation]: A possible concern about that design is that participants would question the reliability of the weak policy. I assess that concern in the supplementary appendix (section 3.2).

@@

I further conducted power estimations for the main effects of the symbols and information manipulations, their conditional effects within each area and the aforementioned interactions. I conducted these estimations after the data collection and analyses, based on the analytical sample size. The estimated detectable (power=0.8) interaction coefficient between areas and the real symbols treatment is 0.27 SDs, and between areas and weak information manipulation is 0.23 SDs. As for the interactions with the relevance manipulation (among Haifa Bay respondents), I estimate that the detectable coefficients of their interactions with the symbols and information conditions are 0.36 and 0.29 SDs, respectively. The power estimations are reported in details in the supplementary appendix (Section 4).

@@

As for the relevance manipulation, data suggests that assignment to relevance priming questions did not increase Haifa Bay respondents' perceptions of personal relevance of the air-pollution policy. In fact, those assigned to the treatment group reported sightly lower level of personal relevance on average, yet these differences are not statistically significant (*t* = `r relevance.exp.t[["statistic"]] %>% round(3)`, *p*=`r round(relevance.exp.t[["p.value"]],3)`). This limitation should be kept in mind when interpreting the results regarding this manipulation. The results of all manipulation and validation checks are available from the supplementary appendix, along with summary statistics and correlation matrix for all variables (sections 3, 5).

@@

[^footnote_raw_means]: Means and SDs of outcome variables across the two areas and the experimental conditions are reported in section XX of the supplementary appendix (section 6).

@@

I further explored the robustness of these findings through additional analyses, reported in the supplementary appendix (section 7). *First*, I replicate the main models while restricting the analyses to the first policy seen by the participants. This enables to rule out the possibility that subjects' responses are affected by the manipulations in the first policy presented to them (a carryover effect), or by their evaluation of the it (assimilation or contrast effects). *Second*, I restrict the analyses to Haifa Bay residents, and compare between those who are more/less likely to perceive the air-pollution as personally relevant, based on two proxies: (a) Subjects city of residence, and its geographical proximity to the polluting industrial area; (b) parents for young children. *Third*, I replicated the analyses on the unfiltered sample. These additional analyses are consistent with the analyses presented in the paper, and do not provide evidence that greater levels of perceived personal relevance attenuates the effect of symbols.

@@

How can we explain the finding that the positive effect of the real symbols was stronger to some extent among Haifa Bay residents? Given that this pattern is also evident for the recycling policy, it is unlikely that it is linked to differences in perceived personal relevance. Rather, I tentatively suggest two other possible explanations for that. First, these patterns could be the result of the higher intercept of center residents' trust in EPM policies, or a ceiling effect. More generally, it could be the case that symbolic elements are more effective in shifting people from low to medium levels of trust (i.e. as in the case of Haifa Bay subjects), than shifting them from medium to high levels (i.e. as in the case of center subjects). In other words, symbols may be more effective in reducing citizens' distrust, then increasing their trust.[^footnote_ceiling_effect] A second explanation could be that EPM symbols attenuated the effect of Haifa Bay residents' negative prior beliefs on their judgments of the policy plans. Accordingly, theoretically there could be an interaction between the emotive effect of symbols and citizens' prior positions embedded in their group affiliation and social identities (in our case, the local identity of Haifa Bay residents). I theorize, tentatively, that the direction of the interaction depends on whether or not the symbolic associations are related to the relevant group identities. When the symbols are not strongly linked to identities of specific groups having strong prior positions on the policy or the organization (as in our case) symbols would attenuate the influence of the prior beliefs on people's judgements. Conversely, when symbolic associations activate group identities and make them more salient, they may enhance the effect of dominant positions embedded in them, and increase polarization between parties [@teodoro_2018]. These theoretical expectations may also have important implications on citizens'  motivated reasoning and biased processing of new information and messages [@taber_2006; @christensen_2018; @baekgaard_2016; @james_2017a]. Hence, future reseasrch may further examine the interactions between different types of symbols and citizens' prior beliefs and motivated reasoning.
